Evolutionary rescue by aneuploidy in tumors exposed to anti-cancer drugs

### 2024-12-22 ResponseLetter.tex

\documentclass[11pt,a4paper]{article}
\usepackage{xcolor}
\usepackage{amssymb,amsmath,amsthm}
\usepackage{natbib}

\renewcommand{\d}{{\rm d}}
\newcommand{\e}{{\rm e}}
\begin{document}
Reviewer $\#1$ :

Reviewer Summary:

In this paper, the authors consider the role of aneuploidy in creating drug-resistant tumor cell lines in cancer. Specifically, they address how aneuploidy contributes to evolutionary rescue of a tumor cell population following treatment with tumor-cell killing drugs. They explore how different types of aneuploidy (stationary, tolerant and partially resistant aneuploid cell populations), time, mutation, rates of change, growth rates of drug resistant cells, pre-treatment aneuploid cell populations, and standing genetic variation, all might contribute, and importantly be exacerbated by aneuploidy, leading to the persistence of tumor cells following chemotherapy that result in a recurrence of tumor, even well past the point of extended remission. The authors use evolutionary models to identify the probability that a tumor could survive and proliferate under different genetic scenarios, rooting their parameters within ranges defined by previous studies, and identify several ways in which aneuploidy may serve a role, directly or in stepping-stone fashion, to facilitate evolutionary rescue of tumor cell populations, especially in smaller tumors often characteristic of primary tumors following resection and secondary tumors following metastasis.

Major Edits:

I have no major edits to recommend on this manuscript and I am comfortable with the authors considering the minor edits and making adjustments as they see fit.

Minor Edits:

1. You do an incredible job of walking the reader through this manuscript, making clear your definitions of important concepts, like evolutionary rescue, along the way. Two places where you didn't do that, but that could really push this paper over the top into perfection territory, are the following: 1) offering your reader an explanation, even if brief, of the multi-type branching method central to your analyses/models, and why it was the right selection for this endeavor among other methods or evolutionary modeling frameworks used in the past or available to you, and, 2) putting something real-world up against the parameter ranges you used to start the models, or just a brief statement about how they were established from the literature you drew upon (i.e., from experimental tests, patient data, other models, etc.). The latter is just in an effort to put something real to the numerical ranges that a reader can relate to before launching into the modeling.

\textcolor{blue}{Thank you for your comment. 
\begin{enumerate}
\item We have added a comment explaining that we use multi-type branching processes because we assume independence between clonal lineages starting from an initial population of $N$ sensitive cells.   change 53 
\item For the sensitive and mutant division rate the paper already states that this parameter is obtain in vivo by Rew and Wilson (2000). I have updated the second paragraph  of "Parameterization" with the fact that Bozic et al. (2013) use patient data in order to obtain the growth rate of sensitive cell. change 54
\end{enumerate}
}

2. Do a search for multi-type branching and make sure you consistently use the hyphen or not.

\textcolor{blue}{Thank you for your comment. We have made the change to use the hyphen in all mentions of "multi-type branching" such that it is consistent with textbooks such as Athreya $\&$ Ney (1972).  change 31}

3. You identify that at least in certain scenarios aneuploidy plays an important role in the evolutionary rescue of tumor cell populations following chemotherapy, at least with one type of drug, and maybe limited to small primary or secondary tumors. But, we never really know the extent that this scenario is applicable. Is it specific to this particular chemotherapy, small tumors, only certain types of cancer? In conjunction with this, there are places in the manuscript where you say things like ". . .for most parameter values aneuploidy will either play no role in evolutionary rescue or will be the main driver of adaptation without requiring any mutation (one step)." Both parts of the quoted sentence stand in opposition to one of the more interesting scenarios found in your modeling, which is that aneuploidy can buy a tumor cell population enough time to accumulate a mutation that leads to evolutionary rescue. Like in point 1, making clear the broader implications of your findings, or the limited scenarios in which they might be applicable, pushes this paper over the top into perfection territory because it roots your reader into something relatable, manageable, or applicable to patient care. If the answer is "I don't know how applicable this finding is or what the broader implications might be," you could then add something to the future research section that talks about how this might be addressed experimentally or theoretically. It's ok to not know for sure, but you shouldn't pass the opportunity to talk about it in limits and future research.

\textcolor{blue}{Thank you for your comment.  We have made an addition to "Future research" that out model can be used on a larger set of tissue-drug combinations to establish whether aneuploidy can play an important role in the evolution of drug resistance. change 55}

4. I would revisit some of your figures to make sure groups/categories are not colored in contradictory ways from one graph to the next (i.e., like "no aneuploidy" being one color on one graph and a different color in another graph or different categories ("no aneuploidy" vs. "aneuploid" vs. "stationary") using the same color in different figures).

\textcolor{blue}{Thank you for your comment.  We have changed Figure 6A to be consistent with the other figures. I have also changed the caption for Figure 6 to be consistent with the new coloring. change 56}

5. The results on detection time associated with tumor size were not intuitive to me on first read. You might want to clarify or rewrite those sentences to make it more clear. It read to me like something about this finding/scenario changed the baseline detection size required to know the tumor cell population was still present and viable.

\textcolor{blue}{Thank you for your comment. We have rewritten the sentence about the mean detection time to explain that for the small and intermediate tumors the mean detection time is approximately  equal to the mean recurrence time. change 48}

6. Table 1 caption doesn't define what the asterisk means.

\textcolor{blue}{Thank you for your comment. We have added the definition of asterisk in the table. change 1}

7. You never tell your reader what ODEs is. . .won't be obvious to everyone. I would just right it out on page 9 where I think is the first time you use it.

\textcolor{blue}{Thank you for your comment. We have added an explanation of the meaning of ODEs to the text in page 9. change 2}

8. Are the tests for standing genetic variation essentially genetic drift? Variation is present and in a tumor cell population wildly fluctuating in size due to treatment you just hit on a "right" variant stochastically when tumor size is small (less cells) that is resistant and it's driven to fixation? If so, using genetic drift in this section is rooting language and another way of describing alternative scenarios/hypotheses for tumor rebound.

\textcolor{blue}{Thank you for your comment. In the context we are discussing, standing genetic variation is generated by the balance between "mutation" (which increases the number of aneuploid cells) and selection (which reduces the number of aneuploid cells). The aneuploid cells are selected against in the pre-tratment regime but once the tumor is treated with anti-cancer drugs they have a fitness advantage over the sensitive cells. If the aneuploids are partially resistant they will grow to fixation if they escape stochastic extinction. However, if the they are stationary of tolerant they will outlast the sensitive cells and increase the porbability of evolutionary rescue.}

9. Do we know if it's reasonable to assume the division rates between cell types (i.e., "sensitive" vs. "mutant") are the same? Maybe ground that assumption into something we know, or just say we are not sure if that is reasonable or not. This relates to point 1 and rooting the reader into these parameter values with something real or being honest about the assumption they are equal.

\textcolor{blue}{Thank you for your comment. It is reasonable to assume that the sensitive and mutant growth rates are similar given that we assume that the mutation makes the cancer cells immune to the chemoterapeutic drugs. change 40}

10. Mid page 7, "decreases" is mis-spelled and I would swap out "very well" with something more descriptive where you use it.

\textcolor{blue}{Thank you for your comment. We have made the change from "decerases" to "decreases". change 27}

Reviewer $\#$2 :

Summary

This is a paper about the likelihood and time for a tumour to evolve drug resistance, via direct mutation or indirectly through aneuploidy. The authors develop a multi-type branching process model to approximate these quantities analytically and to simulate the full behaviour. They also parameterize their approximations and simulations for melanoma, with adaptive trisomies for chromosome 2 and 6. The approximations provide intuition and match simulations well while demonstrating, for instance, that even tolerant (ie, subcritical) aneuploids -- which appears to be the case for melanoma -- can substantially facilitate drug resistance (fig 3) and extend recurrence times (fig 6).

I think this is a very nice application of evolutionary rescue theory. The presentation is clear and the neat approximations provide intuition. Parameterizing for melanoma demonstrates the utility of the theory.

My main concern (elaborated below), is that all of the theoretical results derived here have extremely close analogues in previous evolutionary rescue theory. This substantially reduces the novelty of this paper, in the sense that much of the intuition developed here has been given elsewhere. However, I do think it is useful to derive all of these approximations for a single model that can be parameterized with existing cancer data. What I would like to see, then, is the authors connect their results with existing rescue theory, which would allow them to clarify what is new here and also help develop a largely missing link between the evolutionary rescue and cancer literatures.

$\#$ Major comments

1. The paper is framed as a model about the evolution of drug resistance in tumours via direct mutation or indirectly through aneuploidy. However, the model is much more general than this, it is a model of 1- and 2-step evolutionary rescue that is then parameterized for cancer. Because of this, I would like to see the authors spend more time discussing what is already known about 1- and 2-step rescue, which should also elucidate why another model is needed rather than simply parameterizing a model that already exists. I do think the authors can justify a new model, but I would like to see this developed with a more insightful review of existing models (currently Intro mentions some models exist but doesn't explain why they aren't sufficient) and a comparison of results derived here to those derived previously (currently no rescue papers cited after Intro). In addition to clarifying what is novel here, these comparisons may help connect the currently disparate fields of evolutionary rescue and cancer (see fig 1 in Alexander et al 2014 Evol. Appl.).

In case it helps, my understanding of multi-step rescue theory is that there was an early model for the emergence of epidemics via multiple mutations (Antia et al 2003 Nature, and some follow ups like Yates et al 2006 PRSB and Alexander $\&$ Day 2010 Interface), a very general treatment of emergence and rescue via multiple mutations (Iwasa et al 2004 JTB), a 2-locus 2-allele rescue model (Uecker $\&$ Hermisson 2016 Genetics), and multi-step rescue models on a fitness landscape (Martin et al 2013, Osmond et al 2020), all of which also use branching processes. Many of the results in this paper have been derived in the above studies. For example, eqn 3 is essentially the 1-step rescue probability from Orr $\&$ Unckless 2008 (Am Nat, also their eqn 3) and eqn 4 is essentially the emergence probability from Iwasa et al 2004 (their eqn 3) multiplied by the number of first step mutations that occur during the wildtypes decline. Even more similarly, Eqn 4 is a special case of the results in Osmond et al (2020), who also used the reasoning from fitness valley crossing (Weissman et al 2009) to derive approximations for the probability of 2-step rescue, with the added complication of a distribution of mutation effects. Setting the distribution of fitness effects to a dirac and assuming fast wildtype decline reduces the first two lines of their eqn 8 to the first two lines of eqn 4 (the final line of eqn 4 being the well known case of 1-step rescue). The de novo vs standing variance result (eqn 6) was pointed out in Orr $\&$ Unckless 2008 (right side of page 163) and given in Martin et al 2013 (eqn 3.8). The rescue and recurrence times (eqns 8 and 9) are perhaps the most novel results derived here, but have some close similarities with times derived in Orr $\&$ Unckless 2014 (see eqn 18 and text S2 for rescue time and eqn 23 for recurrence time).

2. I have two general modelling-choice questions whose answers may substantially influence the results. 1) How heritable is a particular aneuploidy? 2) Are aneuploids with a resistance mutation really equivalent to diploids with a resistance mutation? Can the modeling choices made be empirically justified, or can the model (or results) be generalized to include these complications?

$\#$ Minor comments

- line numbers would have been helpful

\textcolor{blue}{Thank you for your comment. We have added line numbers.}

- abstract: first sentence could give more precise definition of evolutionary rescue

\textcolor{blue}{Thank you for your commnet. We have changed the definition of evolutionary rescue such that it highlights the fact that the population must adapt to the environmental change. change 32}

- abstract: 'fitter'$\rightarrow$'more fit'

\textcolor{blue}{Thank you for your comment. We have made the requested change. change 3}

- intro: 'the number of chromosomes and chromosome arms copies'$\rightarrow$'the number of chromosomes and chromosome arm copies'

\textcolor{blue}{Thank you for your comment. We have made the requested change. change 4}

- intro: while there are some similarities, I do not think that evolutionary rescue is mathematically equivalent to the problem of crossing fitness valleys. If that was the case, you wouldn't have to create a new model here and could instead parameterize the model of Weissman et al 2009 for cancer. One difference, for example, is that with fitness valley crossing the population doesn't go extinct and so it is a question about when the valley will be crossed, while in rescue the wildtype is declining to extinction and so the question is largely if rescue will occur. I do agree that the case of two step rescue is similar to fitness valley crossing, but that is a much weaker statement.

\textcolor{blue}{Thank you for your comment. We have changed the text such that it states that evolutionary rescue in two steps is similar to the problem of crossing fitness valley. change 6}

- intro: there is a relatively arbitrary list of rescue papers given at the end of the paragraph, which could be replaced (or supplemented) with synthetic reviews (eg, Alexander et al 2014 Evol Appl, Bell 2017 Annual Reviews)

\textcolor{blue}{Thank you for your comment. We have added the recommend synthetic reviews to the list of rescue papers cited. change 45}

- intro: 'there are three ways for a population to survive ...' - I think this statement would be better placed at the beginning of the paragraph, after which you focus on genetic adaptation

\textcolor{blue}{Thank you for your comment. We have  placed the statement at the beginning of the paragraph. change 28}

- intro: 'most models focus on the probability that at most one mutation rescues the population' - I disagree with this statement. I think the focus is rather on the probability that at least one mutation rescues the population. That is the logic behind the $1-\exp(-Np)$ form: $\exp(-Np)$ is the probability that no mutations rescue the population. In addition to Wilson et al 2017, Osmond $\&$ Coop 2020 (Genetics) also calculate the probability rescue is soft and the expected number of rescuing mutations in a variety of rescue scenarios.

\textcolor{blue}{Thank you for your comment. We have changed the statement to just explain the concepts of soft and hard selective sweeps. Additionaly, we have added a reference to Osmond $\&$ Coop 2020 (Genetics). change 41}

- methods: 'The aneuploid division rate is selected...' - I found the grammar odd here.

\textcolor{blue}{Thank you for your comment. We have changed the statement to be 'The aneuploid division rate is chosen...'.   change 36}

- methods: 'We estimate c=...' - It sounds like you will estimate c from the observed fraction of aneuploid cells but that is not what you describe below.

\textcolor{blue}{Thank you for your comment. We have changed the paragraph to make it clear that we obtaint the fitness cost from the experimental data and then use the formula $f=\frac{\tilde{u}\lambda_s}{c}$ to obtain $f$. change 5}

- methods: is there a justification for not including density-dependence for the sensitives as well? This would hasten their initial decline, which would compound with the reduced initial establishment probability of the mutants. It seems odd they can exert competition on the mutants but don't themselves feel any competition.

\textcolor{blue}{Thank you for your comment. The justification for not including density dependence for sensitive cells is because of the fact that they are under the influence of a powerful stress which causes the population to decline very fast. As a result, density dependent effects will not have an effect on the initial establishment probability of mutants.  As shown in Figure S1, the presence of a carrying capacity does not effect the probability of evolutionary rescue.}

- methods: I got a 404 error for the source code link: https://github.com/yoavram-lab/EvolutionaryRescue

- results: there doesn't appear to be an eqn 1

\textcolor{blue}{Thank you for your comment. We have made changes such that equation 1 appears. change 29}

- results: eqn 3 is an alternate form of the classic rescue probability, eg, $1/N_m^*=u/r*2(s-r)$ in Orr $\&$ Unckless 2008. I think it would be helpful to mention this.

\textcolor{blue}{Thank you for your comment. We have mentioned that equation 2 (the numbers are changes in the updated version) is an alternative form of the formula from Orr $\&$ Unckless 2008. change 33}

- results: eqn 4 is an approximation (fast wildtype decline) of a special case (dirac distribution of mutational effects) in Osmond et al 2020, who used the exact same reasoning from Weissman et al 2009, and is also very similar to eqn 3 in Iwasa et al 2004. I think it would be helpful to mention this.

\textcolor{blue}{Thank you for your comment. We have added a paragraph which explains the similarity of equation 4 (now equation 3) to previous results. change 57}

- results: '$T^*$ decerases'$\rightarrow$'$T^*$ decreases'

\textcolor{blue}{Thank you for your comment. We have made the change from "decerases" to "decreases". change 27}

- results: I thought it was interesting that the second scenario in eqn 5 was the geometric mean of the other two scenarios. Is there some intution for why this is so?

\textcolor{blue}{Thank you for your comment. We have not been able to identify any intuition as to why the second scenario is the geometric mean of the other two scenarios.?}

- results: I thought it might be helpful to give a little more intution about the effect of simultaneously increasing the aneuploid birth and death rates (paragraph starting below eqn 5). The reason why this leads to a decrease in $T^*$ and pushes the system towards the second scenario is that the mutation rate has increased, meaning a lineage doesn't have to persist for as long to almost surely lead to a successful rescue mutant, while the expected persistence time hasn't changed, making it more likely that a rare long-lived aneuploid lineage will lead to rescue.

\textcolor{blue}{Thank you for your comment. We have added the recommended comment about the fact that the reason why $T^*$ decreases, if we increase the aneuploid division and death rates such that $\Delta_a$ remains constant, is because of the fact that more mutants are generated and a lineage doesn't have to persist as long in order to generate a succesful mutant. Change 51 }

- results: I don't get the logic at the end of this paragraph, that your results are consistent with aneuploidy conferring resistance by decreasing the division rate. My understanding is that your result says that, holding aneuploid growth rate constant, increasing aneuploid division rate helps until you reach scenario 2 (effectively stationary/critical aneuploids) but beyond that doesn't help. I don't see how reducing division rate helps rescue.

\textcolor{blue}{Thank you for your comment. In the scenario discribed decreasing the division rate also leads to a decrease in the death rate as the drugs affect the cancer cells during cellular division. We have expanded on the explanation of this scenario in the paper. Change 52}

- results: if $N_a^*$ is similar to the detection threshold, then you predict that any tumour detected is very likely (eg, $>50\%$) to evolve resistance? Does this fit with data on resistance rates?

\textcolor{blue}{Thank you for your comment.  We believe that our model does fit the data on resistance rates given that \citep{handa2022long} state that 68$\%$ of melanoma patients treated with vemurafenib experienced relapse after initial response. change 58}

- results: if aneuploidy may increase the mutation rate and this is easy to implement in the model, why not do it from the start, and reduce to the simpler version $v_a=v$ when needed?

\textcolor{blue}{Thank you for your comment. We started with the simpler model and thought that the more advanced model should be included in Appendix H. Additionally, we do not have biologically realistic values for the aneuploidy induced mutation rate and, as a result, we decided it was better to start with the model for which we have biologically realistic parameters. }

- results: 'cancer cells(' $\rightarrow$ 'cancer cells ('

### 2024-12-22 EvolutionaryRescue1.tex
Remus revision

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{abstract}
Evolutionary rescue happens when a population survives a sudden environmental change that initially causes the population to decline toward extinction.
A prime example of evolutionary rescue is the ability of cancer to survive exposure to treatment. 
One evolutionary mechanism by which a population of cancer cells can adapt to chemotherapy is aneuploidy. 
Aneuploid cancer cells can be fitter in an environment altered by anti-cancer drugs, e.g., because aneuploidy disrupts the pathways usually targeted by the drugs. 
Indeed, aneuploidy is highly prevalent in tumors, and some anti-cancer drugs fight cancer by increasing chromosomal instability.
Here, we model the impact of aneuploidy on the fate of a population of cancer cells. We use multi-type branching processes to approximate the probability that a tumor survives drug treatment as a function of the initial tumor size, the rates at which aneuploidy and other beneficial mutations occur, and the growth rates of the drug-sensitive and drug-resistant cells. Additionally, we investigate the effect of the pre-existent aneuploid cells on the probability of evolutionary rescue.  Finally, we estimate the tumor's mean recurrence time to revert to its initial size following treatment and evolutionary rescue.
We propose that aneuploidy can play an essential role in the relapse of smaller secondary tumors.
\end{abstract}

@@

\paragraph{Aneuploidy in cancer.}  
Each year, approximately 10 million people die from cancer \citep{kocarnik2022cancer}. Understanding the factors that contribute to the failure of interventions is of great importance.
One suggested factor is aneuploidy, in which cells are characterized by an imbalanced karyotype and alterations in the number of chromosomes \citep{schukken2018cin}. 
Aneuploidy is caused by chromosomal instability and mis-segregation of chromosomes during mitosis. Importantly, changes in the number of chromosomes and chromosome arms copies allow cancer cells to survive under stressful conditions such as drug therapy \citep{lukow2021chromosomal,rutledge2016selective,ippolito2021gene}.
Indeed, cancer cells are often aneuploid, and aneuploidy is associated with poor patient outcomes \citep{ben2020context,smith2018systematic}.

 \citet{ippolito2021gene} induced aneuploidy in cancer cell lines by exposing them to reversine, a small-molecule inhibitor of the mitotic kinase Mpsi1, and then to anti-cancer drugs such as vemurafenib. Reversine-treated cells had a higher proliferation rate following drug exposure than sensitive cancer cells, due to selection of specific beneficial karyotypes.
Similarly, \citet{lukow2021chromosomal} induced aneuploidy in cancer cells and observed that such cells have an advantage compared to sensitive cells during drug treatment despite having lower fitness before the onset of treatment.
One proposed mechanism through which aneuploidy can confer resistance to anti-cancer drugs is by antagonizing cell division, which prevents the drugs from damaging DNA and microtubules \citep{replogle2020aneuploidy}. Other mechanisms are the selection for specific karyotypes that lead to reduced drug metabolism \citep{ippolito2021gene} and elevated levels of DNA damage repair due to higher basal levels of DNA damage \citep{zerbib2023human}.

@@

\paragraph{Evolutionary rescue.} Populations adapted to a specific environment are vulnerable to environmental changes, which might cause the population's extinction. Examples of such environmental changes include climate change, invasive species, and the onset of drug therapies. Adaptation is a race against time as the population size decreases in the new environment~\citep{tanaka2022surviving}. 
\emph{Evolutionary rescue} is the process by which the population acquires a trait that increases fitness in the new environment such that extinction is averted. It is mathematically equivalent to the problem of crossing of fitness valley \citep{weissman2009rate,weissman2010rate}.
There are three potential ways for a population to survive environmental change: migration to a new habitat similar to the one before the onset of environmental change \citep{harsch2014keeping,cobbold2020should,zhou2022range}; adaptation by phenotypic plasticity without genetic modification \citep{carja2019evolutionary,carja2017evolutionary,levien2021non,gunnarsson2020understanding}; and adaptation through genetic modifications, e.g., mutation \citep{gomulkiewicz1995does,uecker2014evolutionary,uecker2016role,uecker2011fixation,orr2014population}.

@@

Most models focus on the probability that at most one mutation rescues the population. How multiple mutations contribute to the survival of the population is less explored, but \citet{wilson2017soft} have shown that evolutionary rescue is significantly enhanced by soft selective sweeps when multiple mutations contribute towards evolutionary rescue. Evolutionary rescue that requires two successive mutations (i.e., two steps) has been investigated by \citet{martin2013probability}, who tested their model's predictions with data from yeast and bacteria experiments. \citet{iwasa2003evolutionary,iwasa2004evolutionary} used multi-type branching process theory to approximate the probability that a population under strong selective pressure can survive extinction via succesive mutations.

Here, we study evolutionary rescue after a sudden environmental change caused by initiating anti-cancer drug treatment. 
We consider a range of effects of aneuploidy, from tolerance to (partial) resistance to the drug \citep{brauner2016distinguishing}.
We estimate the effect of aneuploidy on the tumor's evolutionary rescue probability. When aneuploidy provides drug resistance, it can directly rescue the tumor. However, when aneuploidy provides tolerance to the drug, it may act as an evolutionary ``stepping stone'' or ``springboard'' \citep{Yona2015,martin2013probability,osmond_genetic_2020}, delaying extinction, and thereby allowing the tumor more time to acquire a resistance mutation on top of the aneuploid background. 
As mentioned above, evidence suggests that aneuploidy may be a common strategy for tumor adaptation to drug therapy. Still, it is unknown how often aneuploidy provides tolerance and acts as a stepping stone.
We also estimate the mean time until a tumor cell population reaches its pre-treatment size following drug therapy. 
Given that aneuploidy is present in many tumors before the onset of therapy \citep{lukow2021chromosomal,ben2020context}, we also consider the effect of pre-treatment standing genetic variation on the evolutionary dynamics. Additionally, we are interested in the timescale of evolutionary rescue and the impact that aneuploidy has on the time necessary for the tumor to overcome drug therapy.

@@

\begin{subequations}
\begin{flalign*}
(+1,0,0)&:\quad \lambda_ss_t\left(1-u-v\right)\quad\left(\text{birth of sensitive cell}\right),\\
(-1,0,0)&:\quad \mu_ss_t\quad\left(\text{death of sensitive cell}\right),\\
(0,+1,0)&:\quad u\lambda_ss_t\quad\left(\text{sensitive cell divides and becomes aneuploid}\right),\\
(0,0,+1)&:\quad v\lambda_ss_t\quad\left(\text{sensitive cell divides and becomes mutant}\right),\\
(0,+1,0)&:\quad \lambda_aa_t\left(1-v\right)\quad\left(\text{birth of aneuploid cell}\right),\\
(0,-1,0)&:\quad \mu_aa_t\quad\left(\text{death of aneuploid cell}\right),\\
(0,0,+1)&:\quad v\lambda_aa_t\quad\left(\text{aneuploid cell divides and becomes mutant}\right),\\
(0,0,+1)&:\quad \lambda_mm_t\quad\left(\text{birth of mutant cell}\right),\\
(0,0,-1)&:\quad \mu_mm_t\quad\left(\text{death of mutant cell}\right).
\end{flalign*}
\end{subequations}
For the remaining of this paper, we assume that the division rates of sensitive and aneuploid cells can be written as $\lambda_ss_t\left(1-u-v\right)\approx \lambda_ss_t$ and $\lambda_aa_t\left(1-v\right)\approx\lambda_aa_t$ because $u,v\ll1$ (\Cref{table1}).
Each iteration of the simulation loop starts by computing the rates $\nu_k$ of each event $k$.
We then draw the time until the next event, $\Delta t$, from an exponential distribution whose rate parameter is the sum of the rates of all events, such that $\Delta t \sim \text{Exp}(\sum_j \nu_j)$.
Then, we randomly determine which event occurred, where the probability for event $k$ is $p_k = \nu_k/\sum_j \nu_j$.
Finally, we update the number of cells of each genotype according to the event that occurred and update the time from $t$ to $t+\Delta t$.
We repeat these iterations until either the population becomes extinct (the number of cells of all genotypes is zero) or the number of mutant cells is high enough so that their extinction probability is $<0.1\%$, that is, until
\begin{equation*}
m_t > \left\lfloor\frac{3\log{10}}{\log{\left(\lambda_m / \mu_m\right)}}\right\rfloor + 1 ,
\end{equation*}
which we obtain by solving $1-(1-p_m)^{m_t}=0.999$ for $m_t$ with $p_m=\Delta_m/\lambda_m$ as the probability that a single mutant escapes stochastic extinction (Appendix A).

@@

To analyze evolutionary rescue in our model, we use the framework of \emph{multitype branching processes} \citep{harris1963theory, weissman2009rate}. 
This allows us to find explicit expressions for the \emph{survival probability}: the probability that a lineage descended from a single cell does not become extinct.

>>>>

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{abstract}
Evolutionary rescue happens when a population is able to adapt a sudden environmental change that initially causes the population to decline toward extinction. % change 32
A prime example of evolutionary rescue is the ability of cancer to survive exposure to treatment. 
One evolutionary mechanism by which a population of cancer cells can adapt to chemotherapy is aneuploidy. 
Aneuploid cancer cells can be more fit in an environment altered by anti-cancer drugs, e.g., because aneuploidy disrupts the pathways usually targeted by the drugs. %change 3 
Indeed, aneuploidy is highly prevalent in tumors, and some anti-cancer drugs fight cancer by increasing chromosomal instability.
Here, we model the impact of aneuploidy on the fate of a population of cancer cells. We use multi-type branching processes to approximate the probability that a tumor survives drug treatment as a function of the initial tumor size, the rates at which aneuploidy and other beneficial mutations occur, and the growth rates of the drug-sensitive and drug-resistant cells. Additionally, we investigate the effect of the pre-existent aneuploid cells on the probability of evolutionary rescue.  Finally, we estimate the tumor's mean recurrence time to revert to its initial size following treatment and evolutionary rescue.
We propose that aneuploidy can play an essential role in the relapse of smaller secondary tumors.
\end{abstract}

@@

\paragraph{Aneuploidy in cancer.}  
Each year, approximately 10 million people die from cancer globally \citep{kocarnik2022cancer}. Understanding the factors that contribute to the failure of interventions is of great importance. %change 30
One suggested factor is aneuploidy, in which cells are characterized by an imbalanced karyotype and alterations in the number of chromosomes \citep{schukken2018cin}. 
Aneuploidy is caused by chromosomal instability and mis-segregation of chromosomes during mitosis. Importantly, changes in the number of chromosomes and chromosome arm copies allow cancer cells to survive under stressful conditions such as drug therapy \citep{lukow2021chromosomal,rutledge2016selective,ippolito2021gene}. %change 4
Indeed, cancer cells are often aneuploid, and aneuploidy is associated with poor patient outcomes \citep{ben2020context,smith2018systematic}.

 \citet{ippolito2021gene} induced aneuploidy in cancer cell lines by exposing them to reversine, a small-molecule inhibitor of the mitotic kinase Mpsi1, and then to anti-cancer drugs such as vemurafenib. Reversine-treated cells had a higher proliferation rate following drug exposure than sensitive cancer cells, due to selection of specific beneficial karyotypes.
Similarly, \citet{lukow2021chromosomal} induced aneuploidy in cancer cells and observed that such cells have an advantage compared to sensitive cells during drug treatment despite having lower fitness before the onset of treatment.
One proposed mechanism through which aneuploidy can confer resistance to anti-cancer drugs is by extending the length of the cell cycle, which prevents the drugs from damaging DNA and microtubules \citep{replogle2020aneuploidy}. Other mechanisms are the selection for specific karyotypes that lead to reduced drug metabolism \citep{ippolito2021gene} and elevated levels of DNA damage repair due to higher basal levels of DNA damage \citep{zerbib2023human}. %change 34

@@

\paragraph{Evolutionary rescue.} Populations adapted to a specific environment are vulnerable to environmental changes, which might cause the population's extinction. Examples of such environmental changes include climate change, invasive species, and the onset of drug therapies. Adaptation is a race against time as the population size decreases in the new environment~\citep{tanaka2022surviving}. 
\emph{Evolutionary rescue} is the process by which the population acquires a trait that increases fitness in the new environment such that extinction is averted. Evolutionary rescue which require two steps is similar to the problem of crossing of fitness valley with the added difficulty that the population can go extinct~\citep{weissman2009rate,weissman2010rate}. %change 6

There are three potential ways for a population to survive environmental change: migration to a new habitat similar to the one before the onset of environmental change \citep{harsch2014keeping,cobbold2020should,zhou2022range}; adaptation by phenotypic plasticity without genetic modification \citep{carja2019evolutionary,carja2017evolutionary,levien2021non,gunnarsson2020understanding}; and adaptation through genetic modifications, e.g., mutation \citep{gomulkiewicz1995does,uecker2014evolutionary,uecker2016role,uecker2011fixation,orr2014population,alexander2014evolutionary,bell2017evolutionary}. %change 28, 45

@@

Evolutionary rescue occurs when a beneficial mutation appears in a single ancestor, that is, hard selective sweeps, or when multiple mutations appear independently in different ancestors, that is soft selective sweeps. \citet{wilson2017soft} have shown that evolutionary rescue is significantly enhanced by soft selective sweeps when multiple mutations contribute towards evolutionary rescue. Additionally, \citet{osmond2020genetic} calculate the probability that rescue is soft and the expected number of rescue mutations in a variety of scenarios. Evolutionary rescue that requires two successive mutations (i.e., two steps) has been investigated by \citet{martin2013probability}, who tested their model's predictions with data from yeast and bacteria experiments. \citet{iwasa2003evolutionary,iwasa2004evolutionary} used multi-type branching process theory to approximate the probability that a population under strong selective pressure can survive extinction via succesive mutations. %change 41

Here, we study evolutionary rescue after a sudden environmental change caused by initiating anti-cancer drug treatment. 
We consider a range of effects of aneuploidy, from tolerance to (partial) resistance to the drug \citep{brauner2016distinguishing} (we note that we use tolerant to refer to aneuploids which have negative growth rates, which is different from other references in the literature such as \citet{berman2020drug}). %change 47
We estimate the effect of aneuploidy on the tumor's evolutionary rescue probability. When aneuploidy provides drug resistance, it can directly rescue the tumor. However, when aneuploidy provides tolerance to the drug, it may act as an evolutionary ``stepping stone'' or ``springboard'' \citep{Yona2015,martin2013probability,osmond_genetic_2020}, delaying extinction, and thereby allowing the tumor more time to acquire a resistance mutation on top of the aneuploid background. 
As mentioned above, evidence suggests that aneuploidy may be a common strategy for tumor adaptation to drug therapy. Still, it is unknown how often aneuploidy provides tolerance and acts as a stepping stone.
We also estimate the mean time until a tumor cell population reaches its pre-treatment size following drug therapy. 
Given that aneuploidy is present in many tumors before the onset of therapy \citep{lukow2021chromosomal,ben2020context}, we also consider the effect of pre-treatment standing genetic variation on the evolutionary dynamics. Additionally, we are interested in the timescale of evolutionary rescue and the impact that aneuploidy has on the time necessary for the tumor to overcome drug therapy.

@@

\begin{align*}
(+1,0,0)&:\quad \lambda_ss_t\left(1-u-v\right)\quad\left(\text{birth of sensitive cell}\right),\\
(-1,0,0)&:\quad \mu_ss_t\quad\left(\text{death of sensitive cell}\right),\\
(0,+1,0)&:\quad u\lambda_ss_t\quad\left(\text{sensitive cell divides and becomes aneuploid}\right),\\
(0,0,+1)&:\quad v\lambda_ss_t\quad\left(\text{sensitive cell divides and becomes mutant}\right),\\
(0,+1,0)&:\quad \lambda_aa_t\left(1-v\right)\quad\left(\text{birth of aneuploid cell}\right),\\
(0,-1,0)&:\quad \mu_aa_t\quad\left(\text{death of aneuploid cell}\right),\\
(0,0,+1)&:\quad v\lambda_aa_t\quad\left(\text{aneuploid cell divides and becomes mutant}\right),\\
(0,0,+1)&:\quad \lambda_mm_t\quad\left(\text{birth of mutant cell}\right),\\
(0,0,-1)&:\quad \mu_mm_t\quad\left(\text{death of mutant cell}\right).
\end{align*}
For the remaining of this paper, we assume that the division rates of sensitive and aneuploid cells can be written as $\lambda_ss_t\left(1-u-v\right)\approx \lambda_ss_t$ and $\lambda_aa_t\left(1-v\right)\approx\lambda_aa_t$ because $u,v\ll1$ (\Cref{table1}).
Each iteration of the simulation loop starts by computing the rates $\nu_k$ of each event $k$.
We then draw the time until the next event, $\Delta t$, from an exponential distribution whose rate parameter is the sum of the rates of all events, such that $\Delta t \sim \text{Exp}(\sum_j \nu_j)$.
Then, we randomly determine which event occurred, where the probability for event $k$ is $p_k = \nu_k/\sum_j \nu_j$.
Finally, we update the number of cells of each genotype according to the event that occurred and update the time from $t$ to $t+\Delta t$.
We repeat these iterations until either the population becomes extinct (the number of cells of all genotypes is zero) or the number of mutant cells is high enough so that their extinction probability is $<0.1\%$, that is, until
\begin{equation*}
m_t > \left\lfloor\frac{3\log{10}}{\log{\left(\lambda_m / \mu_m\right)}}\right\rfloor + 1 ,
\end{equation*}
which we obtain by solving $1-(1-p_m)^{m_t}=0.999$ for $m_t$ with $p_m=\Delta_m/\lambda_m$ as the probability that a single mutant escapes stochastic extinction (Appendix A).

@@

To analyze evolutionary rescue in our model, we use the framework of \emph{multi-type branching processes} \citep{harris1963theory, weissman2009rate}. % change 31
This allows us to find explicit expressions for the \emph{survival probability}: the probability that a lineage descended from a single cell does not become extinct.
